---
title: "Michelson Speed-of-light Measurements"
author: "An Grocki"
date: 2022-02-08
output:
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: When studying physical problems, there is an important
distinction between *error* and *uncertainty*. The primary purpose of
this challenge is to dip our toes into these factors by analyzing a real
dataset.

*Reading*: [Experimental Determination of the Velocity of
Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
(Optional)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------|------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to
measure the speed of light. His approach was a development upon the
method of Foucault[3], and resulted in a new estimate of
$v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very
close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this
challenge, you will analyze Michelson's original data, and explore some
of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code
chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the
value `LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition)
in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse
```

*Data dictionary*:

-   `Date`: Date of measurement
-   `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair,
    1 = poor
-   `Temp`: Ambient temperature (Fahrenheit)
-   `Velocity`: Measured speed of light (km / s)

### **q1** Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n   | MeanVelocity |
|--------------|-----|--------------|
| 3            | 46  | 299860       |
| 2            | 39  | 299860       |
| 1            | 15  | 299810       |

```{r q1-task}
## TODO: Compute summaries
df_q1 <- df_michelson %>% 
  group_by(Distinctness) %>% 
  summarize(
    n = n(),
    MeanVelocity = as.integer(mean(Velocity))
  )
df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()

```

**Observations**: - Write your observations here! - Images with a
distinctness measurement of 3(good) were most common then 2(fair). A
distinctness of 1(poor) was significantly less common than 3 and 2. -
Why might your table differ from Michelson's? -My table might differ
from Michelson's due to the rounding error. Michelson's table was in
integers, but my initial MeanVelocity had decimals, therefore I
converted it to a integer which could cause some variation since the
only difference between the our tables are the MeanVelocity value. We
most likely used different rounding techniques.

The `Velocity` values in the dataset are the speed of light *in air*;
Michelson introduced a couple of adjustments to estimate the speed of
light in a vacuum. In total, he added $+92$ km/s to his mean estimate
for `VelocityVacuum` (from Michelson (1880), pg. 141). While the
following isn't fully rigorous ($+92$ km/s is based on the mean
temperature), we'll simply apply this correction to all the observations
in the dataset.

### **q2** Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- df_michelson %>% 
  mutate(VelocityVacuum = Velocity + 92)
df_q2
```

As part of his study, Michelson assessed the various potential sources
of error, and provided his best-guess for the error in his
speed-of-light estimate. These values are provided in
`LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently,
Michelson believed the true value of the speed-of-light probably lay
between `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and
`LIGHTSPEED_MICHELSON + LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

-   **Error** is the difference between a true value and an estimate of
    that value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
-   **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does
not know the error. The best they can do is quantify their degree of
uncertainty. We will learn some means of quantifying uncertainty in this
class, but for many real problems uncertainty includes some amount of
human judgment.[2]

### **q3** Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
## Your code here!
# Greater than min estimate 
LIGHTSPEED_VACUUM >= LIGHTSPEED_MICHELSON - LIGHTSPEED_PM
# Greater than max estimate 
LIGHTSPEED_VACUUM <= LIGHTSPEED_MICHELSON + LIGHTSPEED_PM
# within range 
LIGHTSPEED_VACUUM >= LIGHTSPEED_MICHELSON - LIGHTSPEED_PM &
LIGHTSPEED_VACUUM <= LIGHTSPEED_MICHELSON + LIGHTSPEED_PM

#True error
ERROR <- LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON
PERCENT_ERROR <- ERROR/LIGHTSPEED_VACUUM * 100
RATIO <-ERROR/LIGHTSPEED_PM
RATIO
ERROR
PERCENT_ERROR
LIGHTSPEED_PM

```

**Observations**: - Is Michelson's estimate of the error (his
uncertainty) greater or less than the true error? - Michelson's estimate
of the error was less than the true error because the true value does
not fall within range of his estimates. Although the speed is within the
condition that it is less than `LIGHTSPEED_MICHELSON + LIGHTSPEED_PM` ,
it does not meet the condition that is greater than the minim estimate
of `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM`. - Make a quantitative
comparison between Michelson's uncertainty and his error. - Michelson's
true error was -151.542, meaning his estimate was 151.542 larger than
the true value. His estimate, 51, was lower than the 151.542. Although
his uncertainty was lower ( a smaller estimated error) than his true
error, his percent error of 0.05 is extremely low which helps put the
numbers into perspective. The true error was 2.97x, almost 3x, bigger
than Michelson's estimate.

The following plot shows all of Michelson's data as a [control
chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot
is common in manufacturing, where it is used to help determine if a
manufacturing process is under [statistical
control](https://en.wikipedia.org/wiki/Statistical_process_control).
Each dot is one of Michelson's measurements, and the grey line connects
the mean taken for each day. The same plot also shows simulated data
using a probability model. We'll get into statistics later in the
course; for now, let's focus on understanding what real and simulated
data tend to look like.

### **q4** Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**: Similarities -The majority of both data sets fall
within the range of around 299900 and 300000 as shown with the dotted
lines. Both data sets also have points outside Michelson's range of the
speed of light indicated by the dashed black lines and collect multiple
data samples in one day. Differences - The real data seems to have more
variation at the begining of June where the simulated data seems to have
an even amount of variation throughout the different dates. In addition,
the real data seems to have more variation in total than the simulated
graph.

### **q5** You have access to a few other variables. Construct a few visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?

```{r}
df_q2 %>%
  ggplot(aes(Temp, VelocityVacuum)) +
  geom_point(aes(x = Temp, y = VelocityVacuum)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dotted",
    color = "red"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  )
```

```{r}
df_q2 %>%
  ggplot(aes(Date, VelocityVacuum)) +
  geom_point(aes(x = Date, y = VelocityVacuum, color = Distinctness )) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dotted",
    color = "red"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  )
```

**Observations**:

-   As the temperature increases, the velocity seems to have more
    variation although there is not a clear trend of the velocity
    increasing to decreasing.
-   The distinctness does not seem to affect the difference between the
    Michelson velocity and the true velocity.
-   Around June 16, the distinctness of some of the images was poor.
-   The majority of the data was measured to be higher than the true
    velocity value, and generally not significantly impacted by the
    different variables. I believe the error was in the procedure that
    resulted in higher measured velocities.
-   No, there are no patterns in the data that might explain the
    difference between Michelson's estimate and \`LIGHTSPEED_VACUMN\`.

## Bibliography

-   [1] Michelson, [Experimental Determination of the Velocity of
    Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
    (1880) 
-   [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical
    Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf)
    (1986) 
-   [3] BYU video about a [Fizeau-Foucault
    apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to
    what Michelson used.
