---
title: "Gapminder"
author: "An Grocki"
date: 2023-02-23 
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
summary(gapminder)
```

**Observations**:

-   Write all variable names here
    -   country

    -   continent

    -   year

    -   lifeExp

    -   pop

    -   gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% 
  pull(year) %>% 
  max()
year_min <- gapminder %>% 
  pull(year) %>% 
  min()



```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
#Scatter Plot
year_min
gapminder %>% 
  filter(year == year_min) %>% 
  ggplot() + 
  geom_point(
    mapping = aes(y = gdpPercap, x = continent)
  )
#Box Plot
gapminder %>% 
  filter(year == year_min) %>% 
  ggplot() + 
  geom_boxplot(
    mapping = aes(y = gdpPercap, x = continent)
  ) 
#Box Plot with Log Scale
gapminder %>% 
  filter(year == year_min) %>% 
  ggplot() + 
  geom_boxplot(
    mapping = aes(y = gdpPercap, x = continent)
  ) +
 scale_y_log10()
#Box Plot after removing an outlier 
gapminder %>% 
  filter(year == year_min, gdpPercap < 90000) %>% 
  ggplot() + 
  geom_boxplot(
    mapping = aes(y = gdpPercap, x = continent)
  ) 
```

**Observations**:

-   Observations of gdpPercap vs continent in 1952
-   Plot 1: Scatter Plot
    -   Oceania has significantly fewer data points than the other continents.

    -   The majority of gdpPercap are clustered below 15,000.

    -   Africa appears to have the lowest spread of gdpPercap.

    -   Asia has an outlier that significantly greater than other data points.
-   Plot 2: Box Plot
    -   Asia has an outlier that significantly greater than other data points.

    -   Oceania gdpPercap were clustered together and had the highest median.

    -   America, Africa, Asia and Europe have outliers signified with black dots that are.
-   Plot 3: Box Plot with Log Scale
    -   America has two outliers and Asia has 1, but the previous outliers shown without log scale no longer appear.

    -   Excluding Asia's outlier, Asia seems to have the largest spread of data.
-   Plot 4: Box Plot with Removing an Outlier
    -   Based on the median value of each countries gdpPercap, Oceania was highest, then Europe, Americas, Asia, and Africa.

    -   Based on the box plots, Africa had the lowest spread of 50% of data.

    -   Europe had the largest spread of data.

**Difficulties & Approaches**:

-   I initially created a scatter plot, but due to the large scales and cluster of points at the bottom, it was hard to read the data.
-   I switched to using a box plot to understand the distribution of data, but encountered a similar problem to the scatter plot because of the outlier in Asia increased the y scale and made the box plots hard to read.
-   To solve the problem, I switched to using a log scale which helped me gain a broad understanding of the data without removing any pieces. Personally for me, log scales are difficult to read and understand the details, but log scales help show the spread of data without removing outleirs.
-   To understand the graph more clearly, I removed the outlier by filtering gdp by values below 90,000 excluding an outlier. I thought the data was the most clear to read in this form
-   I am defining an outlier as a data point that differs significantly from other observations. In a box plot, an outlier is represented as a black dot.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
# Box Plot after removing an outlier  
outliers <- gapminder %>% 
  filter(
    year == year_min,
    gdpPercap > 11000,
    continent != "Oceania"
  ) %>% 
  arrange(desc(gdpPercap))
outliers
```

**Observations**:

-   Identify the outlier countries from q2
    -   Kuwait, Switzerland, United States, and Canada.
    -   Outliers below 11,000 gdpPercap are not found.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>% 
  filter(year == year_min | year == year_max) %>% 
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "Switzerland", "United States", "Canada")),
    mapping = aes(color = country),
    size = 2
  ) +
  facet_wrap(~ year)

gapminder %>% 
  filter(year == year_min | year == year_max) %>% 
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +  
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "Switzerland", "United States", "Canada")),
    mapping = aes(color = country),
    size = 2
  ) +
  facet_wrap(~ year) + 
  scale_y_log10()
  
```

**Observations**:

-   From 1952 to 2007, United States, Switzerland, and Canada gpd went up while Kuwait significantly decreased.
-   In 2007, United States and Canada are still outliers, but Kuwait and and Switzerland are no longer outliers.
-   From 1952 to 2007, the interquartile range of every continent increased.
-   Outliers are identified based on box plot as black dots.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## Find outliers I missed earlier.
## Africa
gapminder %>% 
  filter((year == year_min | year == year_max),
         continent == "Africa") %>% 
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +  geom_point(
    data = . %>% filter(country %in% c("South Africa", "Gabon", "Angola", "Reunion", "Botswana", "Equatorial Guinea", "Libya", "Mauritius")),
    mapping = aes(color = country),
    size = 2
  ) +
  facet_wrap(~ year) 

# Filtering outliers 
Africa_outliers_1952 <- gapminder %>% 
  filter(
    year == year_min,
    gdpPercap > 2700,
    continent == "Africa"
  ) %>%
   arrange(desc(gdpPercap))
Africa_outliers_1952
Africa_outliers_2007 <- gapminder %>% 
  filter(
    year == year_max,
    gdpPercap > 7600,
    continent == "Africa"
  ) %>%
   arrange(desc(gdpPercap))
Africa_outliers_2007
```

-   Based on my initial box graph of Africa in 1952 and 2007, I filtered and labeled the outliers from both 1952 and 2007
-   In 1952, South Africa, Gabon, Angola, and Reunion were outliers that were higher than upper range.
-   In 2007, Gabon, Botswana, Equatorial Guinea, Libya, Mauritius, South Africa and Reunion were outliers that were higher than the upper range.
-   Although the interquartile range increased, the median only increased slightly.
-   Equatorial Guinea increased significantly from the lower range to an outlier above the higher range.
-   Most countries there were outliers in either 1952 or 2007 increased significantly except for Angola.
-   Questions:
    -   What is the historical context on why Equatorial Guinea increased?

    -   How do the outliers found in the earliest and latest data change in the different years?

```{r q5-task2}
## TASK: Your second graph
## See how the outliers from 1952 and 2007 increase or decrease over the years 
## Africa 
gapminder %>% 
  filter(continent == "Africa") %>% 
  ggplot(aes(year, gdpPercap)) +
  geom_point() +  
  geom_point(
    data = . %>% filter(country %in% c("South Africa", "Gabon", "Angola", "Reunion", "Botswana", "Equatorial Guinea", "Libya", "Mauritius")),
    mapping = aes(color = country),
    size = 2
  ) +
  geom_line(
    data = . %>% filter(country %in% c("South Africa", "Gabon", "Angola", "Reunion", "Botswana", "Equatorial Guinea", "Libya", "Mauritius")),
    mapping = aes(color = country),
  ) 
  

```

-   Following the paths of outliers found in either 1952 or 2007, the gdp changes dramatically between different countries.
-   Mauritius, Reunion, Botswana, and South Africa had consistent and steady growth from 1952 to 2007.
-   Equatorial Guinea remained around the same gdp until around 1992 when it started increasing rapdily.
-   Gabon and Libya increased rapidly after 1960 then dropped significantly from 1975 to around 1985.
-   Questions
    -   What happened after 1975? (When both Gabon and Libya dramatically increased then dropped in gdp)

    -   Does population or life expectancy follow similar trends?

    -   Does countries with a higher gdp have a higher life expectancy?

```{r q5-task3}
## See if the life expectancy follows a similar trend 
#Afria 
gapminder %>% 
  filter(continent == "Africa") %>% 
  ggplot(aes(year, lifeExp)) +
  geom_point() +  
  geom_point(
    data = . %>% filter(country %in% c("South Africa", "Gabon", "Angola", "Reunion", "Botswana", "Equatorial Guinea", "Libya", "Mauritius","Rwanda")),
    mapping = aes(color = country),
    size = 2
  ) +
  geom_line(
    data = . %>% filter(country %in% c("South Africa", "Gabon", "Angola", "Reunion", "Botswana", "Equatorial Guinea", "Libya", "Mauritius", "Rwanda")),
    mapping = aes(color = country),
  ) 

#outlier 
lifeExp_outlier <- gapminder %>% 
  filter(
    lifeExp < 30,
    continent == "Africa"
  ) 

lifeExp_outlier


```

-   Life Expectancy does not follow a similar trend to GDP.
-   In general life expectancy slowly increases as years increase.
-   Following the paths of outlier countries, the trend is also found where as years increases, life expectancy also increases.
    -   Although the data generally follows this trend over time, there are 4 counter examples for certain lengths of time. From around 1990 to 2002, Gabon, Botswana, and South Africa decrease in life expectancy as opposed to increasing. From 1980 to around 1990, Rwanda also decreases in life expectancy.
-   Questions
    -   What happened to the outlier life expectancy that is below 25 years?

        -   The country was Rwanda in 1992 with a life expectancy of 23.559. In 1990s, Rwanda had a Civil War that lead into a genocide(I am very sad now).

        -   I added the country Rwanda to the list of countries I'm tracking due to interest in the outlier.

    -   What happened in Africa around 1990 where multiple life expectancy of different countries dropped?

    -   What other factors impact the broad range of life expectancy in different years.

    -   Does life expectancy taper off and when?
