---
title: "SAT and College Grades"
author: "An Grocki"
date: 2022-04-19
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: How do we apply hypothesis testing to investigating data? In
this challenge you'll practice using hypothesis testing tools to make
sense of a dataset.

*Reading*: - [Harvard Study Says SATs Should Be Optional: Here's
Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why)
(Optional); easy-to-read news article on colleges going SAT-free -
[Norm-Referenced Tests and Race-Blind
Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university)
(Optional); technical report on relationship between the SAT/ACT and
non-academic factors

*Credit*: This is based on a [case
study](http://onlinestatbook.com/2/case_studies/sat.html) originally
prepared by Emily Zitek, with data collected through the research of
Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
```

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

*Background*: Every year about 2 million students take the Scholastic
Aptitude Test (SAT). The exam is
[controversial](http://www.nea.org/home/73288.htm) but [extremely
consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html).
There are many claims about the SAT, but we're going to look at just
one: Is the SAT predictive of scholastic performance in college? It
turns out this is a fairly complicated question to assess---we'll get an
introduction to some of the complexities.

# Obtain the Data

<!-- -------------------------------------------------- -->

### **q1** Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <- read_excel("data/sat.xls")

## TODO: Do your "first checks"
summary(df_sat)
glimpse(df_sat)

```

**Observations**:

-   Fill in the following "data dictionary"

| Column     | Meaning                                |
|------------|----------------------------------------|
| `high_GPA` | High school grade point average        |
| `math_SAT` | Math SAT score                         |
| `verb_SAT` | Verbal SAT score                       |
| `comp_GPA` | Computer science grade point average   |
| `univ_GPA` | Overall university grade point average |

-   What information do we have about these students?
    -   We have their high school GPA, math SAT, verbal SAT, computer
        science and university GPA.
-   What kinds of information *do we not have* about these students?
    -   Their background and which school they go to in both high school
        and university.
-   Based on these missing variables, what possible effects could be
    present in the data that we would have *no way of detecting*?
    -   How their background or school affected their SAT score or GPA.
        We don't know if the data set is biased towards one factor or
        demographic.

# Analysis with Hypothesis Testing

<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the
first based on hypothesis testing of correlation coefficients, and the
second based on fitting a regression model and interpreting the
regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)
```

## View 1: Correlations

<!-- ----------------------------------------------------------------------- -->

### **q2** Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
## TODO:
df_composite %>% 
  ggplot(aes( x = univ_GPA)) + 
  geom_point(aes(y = high_GPA, color = "high_GPA")) +
  geom_point(aes(y = both_SAT *(4/1600), color = "both_SAT")) +
  scale_y_continuous(
    # Features of the first axis
    name = "High School GPA",
    # Add a second axis and specify its features
    sec.axis = sec_axis( trans=~.*(1600/4), name="Both SAT")
  ) 

```

**Observations**:

-   What relationship do `univ_GPA` and `both_SAT` exhibit?
    -   As SAT score increases, university GPA tends to also increases
-   What relationship do `univ_GPA` and `high_GPA` exhibit?
    -   As high school GPA increases, university GPA also increases

### Hypothesis Testing with a Correlation Coefficient

<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation
coefficient. The idea is to set our null hypothesis to the case where
there is no correlation, and test to see if the data contradict that
perspective. Formally, the null (H0) and alternative (HA) hypotheses
relating to a correlation coefficient between two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the
assumption that `X, Y` are both normally distributed. First, let's check
to see if this assumption looks reasonable for our data.

### **q3** Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
df_composite %>% 
  ggplot(aes(x=both_SAT)) + 
  geom_histogram()
df_composite %>% 
  ggplot(aes(x=high_GPA)) + 
  geom_histogram()
df_composite %>% 
  ggplot(aes(x=univ_GPA)) + 
  geom_histogram()

```

**Observations**:

-   To what extent does `both_SAT` look like a normal distribution?
    -   It does not look like a normal distribution.
-   To what extent does `high_GPA` look like a normal distribution?
    -   It does not look like a normal distribution.
-   To what extent does `univ_GPA` look like a normal distribution?
    -   It does not look like a normal distribution.

Keep in mind your findings as you complete q4.

### **q4** Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA
cor.test(formula = ~ high_GPA + univ_GPA,
                             data = df_composite)
cor.test(formula = ~ both_SAT + univ_GPA,
                             data = df_composite)
```

**Observations**:

-   Which correlations are significantly nonzero?
    -   Both correlations are nonzero because their confidence interval
        does not include 0.
-   Which of `high_GPA` and `both_SAT` seems to be more strongly
    correlated with `univ_GPA`?
    -   High_GPA.
-   How do the results here compare with the visual you created in q2?
    -   Yes, the trend was more clear with high school GPA and
        University GPA. This could also be caused by the way I choose to
        display the information.
-   Based on these results, what can we say about the predictive
    capabilities of both `high_GPA` and `both_SAT` to predict
    `univ_GPA`?
    -   Both high_GPA and both_SAT can somewhat predict univ_GPA since
        they are both correlated.

Finally, let's use the bootstrap to perform the same test using
*different* assumptions.

### **q5** Use the bootstrap to approximate a confidence interval for `corr[high_GPA, univ_GPA`. Compare your results---both the estimate and confidence interval---to your results from q4.

```{r q5-task}
## TODO: Use the bootstrap to compute a confidence interval for corr[high_GPA, univ_GPA]

set.seed(101)
times <- 1000 # Number of bootstrap resamples

df_bootstrap <-
  df_composite %>%
  bootstraps(times = times) %>%
  mutate(df = map(splits, ~ analysis(.x))) %>%
  slice(1:10) %>%
  unnest(df)

cor.test(formula = ~ high_GPA + univ_GPA,
                             data = df_bootstrap)


```

**Observations**:

-   How does your estimate from q5 compare with your estimate from q4?
    -   The correlation is 0.783 correlation which is very similar to
        the estimate in q4, the 0.79.
-   How does your CI from q5 compare with your CI from q4?
    -   The confidence interval in q5 was 0.759- 0.806 while the
        confidence interval in q4 was 0.691- 0.845. The new confidence
        interval is significantly smaller but within the q4 range.

*Aside*: When you use two different approximations to compute the same
quantity and get similar results, that's an *encouraging sign*. Such an
outcome lends a bit more credibility to the results.

## View 2: Modeling

<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study
the relationship among more variables we can instead use a fitted model.
Using a model, we can also help assess whether it is *worthwhile* to
measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model

<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a
model, we can express our hypotheses in terms of the model parameters.
For instance, if we were interested in whether $X$ has an affect on $Y$,
we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on
$Y$. Let's apply this idea to relating the variables `univ_GPA` and
`high_GPA`. Luckily R has built-in tools to construct a confidence
interval on the $\beta$'s in a regression [1]; we'll simply use those
tools rather than do it by hand.

### **q6** Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <-
  df_composite %>%
  lm(
    data = .,               # Data for fit
    formula = univ_GPA ~ both_SAT # Formula for fit
  )

fit_basic %>% tidy()

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

##Checking results
rsquare(fit_basic, df_train)
rsquare(fit_basic, df_validate)
```

**Observations**:

-   What is the confidence interval on the coefficient of `both_SAT`? Is
    this coefficient significantly different from zero?
    -   The confidence interval is 0.0019 -0.00035. The coefficient is
        significantly different from zero because it does not include it
        within the interval.
-   By itself, how well does `both_SAT` predict `univ_GPA`?
    -   Both_SAT does not do well at predicting university GPA.

Remember from `e-model03-interp-warnings` that there are challenges with
interpreting regression coefficients! Let's investigate that idea
further.

### **q7** Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
fit_mid <-
  df_composite %>% 
  lm(formula = univ_GPA ~ high_GPA + both_SAT)

## NOTE: The following computes confidence intervals on regression coefficients
fit_mid %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

#Checking results
rsquare(fit_mid, df_train)
rsquare(fit_mid, df_validate)

```

**Observations**:

-   How well do these models perform, compared to the one you built in
    q6?
    -   The models performs a lot better because the r square values are
        higher.
-   What is the confidence interval on the coefficient of `both_SAT`
    when including `high_GPA` as a predictor?? Is this coefficient
    significantly different from zero?
    -   The confidence interval was -0.000223 - .0018. The coefficient
        could be zero yet it could also different from zero.
-   How do the hypothesis test results compare with the results in q6?
    -   The hypothesis test results are better than the results in q6
        because the r sqaure values are higher in the hypothesis than
        q6.

## Synthesize

<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### **q8** Using the results from all previous q's, answer the following questions.

**Observations**:

-   Between `both_SAT` and `high_GPA`, which single variable would you
    choose to predict `univ_GPA`? Why?
    -   I would choose high_GPA because it is more correlated with
        univ_GPA.
-   Is `both_SAT` an effective predictor of `univ_GPA`? What specific
    pieces of evidence do you have in favor of `both_SAT` being
    effective? What specific pieces of evidence do you have against?
    -   Both_SAT is not an effective predictor of univ_GPA
    -   Effective.
        -   Has a positive correlation with univ_GPA
        -   Potentially could have a slight impact based on the model
    -   Ineffective.
        -   Potentially could have no impact based on model, 0 is within
            range.
        -   High GPA is more effective predictor.

# End Notes

<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more
information see this [Wiki
article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
