---
title: "Massachusetts Highway Stops"
author: "An Grocki"
date: 2022-05-02
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-----------------|--------------------------|-----------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup

<!-- -------------------------------------------------- -->

### **q1** Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "data/ma_statewide_2020_04_01.rds"
df_data <- readRDS(filename)
```

# EDA

<!-- -------------------------------------------------- -->

### **q2** Do your "first checks" on the dataset. What are the basic facts about this dataset?

```{r}
summary(df_data)
```

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### **q3** Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
df_data %>% 
  pull(subject_race) %>% 
  unique()
df_data %>% 
  pull(raw_Race) %>% 
  unique()

```

**Observations**:

-   What are the unique values for `subject_race`?
    -   White, Hispanic, Black, Asian/Pacific Islander, other, NA, unknown
-   What are the unique values for `raw_Race`?
    -   White, Hispanic, Black, Asian or Pacific Islander, Middle Eastern or East Indian(South Asian), American Indian or Alaskan Native, NA, None - for no operator present citations only, A
-   What is the overlap between the two sets?
    -   White, Hispanic, Black, Asian or Pacific islander, NA
-   What is the difference between the two sets?
    -   Other, unknown
    -   Middle Eastern or East Indian(South Asian), American Indian or Alaskan Native, None - for no operator present citations only, A

### **q4** Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.
orginal <- df_data %>% nrow()
matched <- df_data %>% 
  filter(
    is.na(subject_race) & is.na(raw_Race) |
  (subject_race == "white" & raw_Race == "White") |
  (subject_race == "hispanic" & raw_Race == "Hispanic") |
  (subject_race == "black" & raw_Race == "Black") |
  (subject_race == "asian/pacific islander" & raw_Race == "Asian or Pacific Islander")) %>% 
    nrow()
fraction <- matched/orginal
fraction

```

**Observations**

Between the two hypotheses:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

-   It is most plausible that race_Raw is the unprocessed version of subject race since 97 percent of cases match the overlapping values between race_Raw and subject_race.

## Vis

<!-- ------------------------- -->

### **q5** Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

```{r}
# rate by age
df_data %>% 
  drop_na(subject_age) %>% 
  group_by(subject_age) %>%
  summarize(
    num = n(),
    arrest_count = sum(!is.na(arrest_made) & arrest_made == TRUE),
    arrest_rate = arrest_count / num
  ) %>% 
  ggplot(aes(subject_age, arrest_rate)) +
  geom_point()

df_data %>% 
  drop_na(subject_age) %>% 
  group_by(subject_age) %>%
  summarize(
    num = n(),
    arrest_count = sum(!is.na(arrest_made) & arrest_made == TRUE),
    arrest_rate = arrest_count / num
  ) %>% 
  filter(arrest_rate > .05)

```

```{r}
# rate by sex
df_data %>% 
  drop_na(subject_sex) %>% 
  group_by(subject_sex) %>%
  summarize(
    num = n(),
    arrest_count = sum(!is.na(arrest_made) & arrest_made == TRUE),
    arrest_rate = arrest_count / num
  ) %>% 
  ggplot() +
  geom_bar(aes(x = subject_sex, y = arrest_rate), stat = "identity")
```

```{r}
#rate by race 
df_data %>% 
  drop_na(subject_race) %>% 
  group_by(subject_race) %>%
  summarize(
    num = n(),
    arrest_count = sum(!is.na(arrest_made) & arrest_made == TRUE),
    arrest_rate = arrest_count / num
  ) %>% 
  ggplot() +
  geom_bar(aes(x = subject_race, y = arrest_rate), stat = "identity")
```

**Observations**:

-   How does `arrest_rate` tend to vary with `subject_age`?
    -   The arrest rate has a quick spike at 16, before going down then steadily increasing until around 26 before decreasing as the subject's age increases.
-   How does `arrest_rate` tend to vary with `subject_sex`?
    -   Mens arrest rate is almost twice as a females arrest rates.
-   How does `arrest_rate` tend to vary with `subject_race`?
    -   Hispanics have significantly higher arrest rates then other races.
    -   Hispanic \> black \> other \> white \> Asian \> unknown

# Modeling

<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### **q6** Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

-   Which `subject_race` levels are included in fitting the model?
    -   white, black, and Hispanic
-   Which `subject_race` levels have terms in the model?
    -   Hispanic and white

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### **q7** Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
fit_q7  <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ) %>% 
    mutate(subject_race = fct_relevel(subject_race, "white")),
    family = "binomial"
  )
fit_q7 %>% tidy()
fit_q7
```

**Observations**:

-   Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
    -   The highest rate is if the subject's rate is Hispanic and the lowest rate is if the subjects race is white.
-   What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
    -   Racial Basis
    -   Income/class differences
    -   Race of the officer
-   Look at the sent of variables in the dataset; do any of the columns relate to a potential explanation you listed?
    -   Location
    -   Reason for stop
    -   Search conducted

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### **q8** Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q8  <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ) %>% 
    mutate(subject_race = fct_relevel(subject_race, "white")),
    family = "binomial"
  )
fit_q8 %>% tidy()
```

**Observations**:

-   How does controlling for found contraband affect the `subject_race` terms in the model?
    -   The same pattern emerges where the subject's more likely to get arrested if there race if there are Hispanic, then black, then white although at a significantly lower degrees.
-   What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
    -   Finding a contraband tell us that the the police officer searched the car and that finding a contraband is a higher indicator of being arrested than race. The stop does not account for racial basis in the decision to search the car as well as ratios of the cars searched and not searched.

### **q9** Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

-   What factors increase the rates of someone being frisked?(<https://www.nytimes.com/2020/03/02/upshot/stop-and-frisk-bloomberg.html?smid=url-share>)

```{r}
# rate by age
df_data %>% 
  drop_na(subject_age) %>% 
  group_by(subject_age) %>%
  summarize(
    num = n(),
    frisk_count = sum(!is.na(frisk_performed  ) & frisk_performed  == TRUE),
    frisk_rate = frisk_count / num
  ) %>% 
  ggplot(aes(subject_age, frisk_rate)) +
  geom_point()

df_data %>% 
  drop_na(subject_age) %>% 
  group_by(subject_age) %>%
  summarize(
    num = n(),
    frisk_count = sum(!is.na(frisk_performed  ) & frisk_performed  == TRUE),
    frisk_rate = frisk_count / num
  ) %>% 
  filter(frisk_rate > .003)

df_data %>% 
  drop_na(subject_age) %>% 
  group_by(subject_age) %>%
  summarize(
    num = n(),
    frisk_count = sum(!is.na(frisk_performed  ) & frisk_performed  == TRUE),
    frisk_rate = frisk_count / num
  ) %>% 
  filter(frisk_rate == 0)

df_data %>% pull(subject_age) %>% summary()

```

```{r}
# rate by race
df_data %>% 
  drop_na(subject_sex) %>% 
  group_by(subject_sex) %>%
  summarize(
    num = n(),
    frisk_count = sum(!is.na(frisk_performed  ) & frisk_performed  == TRUE),
    frisk_rate = frisk_count / num
  ) %>% 
    ggplot() +
  geom_bar(aes(x = subject_sex, y = frisk_rate), stat = "identity")
```

```{r}
# rate by race
df_data %>% 
  drop_na(subject_race) %>% 
  group_by(subject_race) %>%
  summarize(
    num = n(),
    frisk_count = sum(!is.na(frisk_performed  ) & frisk_performed  == TRUE),
    frisk_rate = frisk_count / num
  ) %>% 
    ggplot() +
  geom_bar(aes(x = subject_race, y = frisk_rate), stat = "identity")

```

```{r}
fit_q9  <-
  glm(
    formula = frisk_performed ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ) %>% 
    mutate(subject_race = fct_relevel(subject_race, "white")),
    family = "binomial"
  )
fit_q9 %>% tidy()
```

**Observations**:

-   Age
    -   From 10-13, you are most likely not going to be frisked, but then it jumps from 14-16 before decreasing until you are 70 when you also tend not to get frisked. There is a odd outlair at 84 years old where the rate is as high as 14-16.
-   Sex
    -   Similar to the results with being arrested, men are almost 3x more likely to be frisked then women.
-   Race
    -   Race is also similar the results from being arrested, although both Hispanic and black are significantly higher.
    -   Hispanic \> Black \> Other \> White \> Asian / Pacific Islander \> unknown
-   Model
    -   Based on the model, being white has the lowest probability of being frisked then women while being Hispanic has the highest probability of being frisked then being black.

<!-- -------------------------------------------------- -->

-   Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
